Quantum Hoare Types

### 2019-12-11 ms-paper/ms-paper.tex
Fix typo, remove review option
Minor text refinements

\documentclass[acmsmall,nonacm,timestamp,review]{acmart}

@@

\begin{abstract}
Quantum Computing is a new and upcoming way to utilize quantum mechanics for computation. It is easy to introduce bugs while programming quantum computers similar to classical computing. Strong static type systems have prevented a lot of bugs from being introduced in the classical paradigm. There is a similar need for well-designed types in the domain of quantum computing. We propose Quantum Hoare Types that let programmers reason statically about certain semantic properties of their quantum programs and further allow a limited form of formal verification.
\end{abstract}

@@

\section{Introduction / Background}
It is difficult to reason about the correctness of quantum programs. Sound static type systems help prevent a huge class of bugs from occurring but since the realm of quantum programming is still new there is not a lot of consensus on what kind of types make the most sense. Further, it is unclear how much they help programmers reason about the semantic properties associated with the quantum algorithms that they are implementing.

@@

\subsection{Quantum Computing}
The basic unit of information in quantum computing is a qubit, that can be represented using a unit vector in a 2-dimensional complex vector space. Using Dirac's bra-ket notation, we can write an arbitrary qubit, $\ket{\psi}$, as a linear combination of the computational basis vectors: $\ket{\psi} = \alpha\ket{0} + \beta\ket{1}$, where $\ket{0}=\begin{bsmallmatrix}
1 \\
0
\end{bsmallmatrix}$ and $\ket{1}=\begin{bsmallmatrix}
0 \\
1
\end{bsmallmatrix}$, and the two constants are complex amplitudes that are normalized using the constraint $|\alpha|^2 +|\beta|^2 = 1$. This means that when measured the probability of the qubit being in the state 0 is $|\alpha|^2$ and that for the state 1 is $|\beta|^2$. To work with multiple qubits, we need to take their tensor product, denoted by $\otimes$. Computation is performed by the application of unitary quantum gates, such as Hadamard (H) or controlled NOT (CX), that can be represented as square matrices. Mathematically, a gate application is equivalent to multiplying the matrix representing the gate with the state vector of the qubit(s).

This basic unit of information is not the only reason why quantum computing is fundamentally different from classical computing---certain quantum mechanical phenomenon are highly counterintuitive! One such phenomenon is \textit{quantum entanglement} which, at its most basic level, involves a correlation between two qubits such that when one is measured the outcome obtained necessarily influences the measurement outcome of the other qubit (even if the two qubits are far apart). We will see how this can be utilized for computation in the quantum teleportation algorithm that we describe and program in \S\ref{telep}. We refer the reader to the standard textbook~\cite{nielsen_quantum_2010} or Nielsen's excellent series of essays~\cite{matuschak2019} for more background.

@@

\subsection{Example: Quantum Teleportation}\label{telep}
Teleportation is a simple quantum algorithm that is used to communicate a quantum state using two classical bits and a shared entangled pair of qubits. Alice is trying to send a message encoded in the qubit to Bob. There are three broad steps in the protocol: 1) preparation, ie, entanglement creation between two qubits to be shared between Alice and Bob in advance; 2) sending the message by Alice; and, 3) reconstruction of the message by Bob. Program and circuit follow:
\begin{figure}[h]
	\begin{subfigure}{3in}
		% 	a       ← init0 ();
		%     b       ← init0 ();
		\begin{lstlisting}
		teleport : Π\kp:qbit.{q↦qbit\kp /\ a↦qbit|0⟩ /\ b↦qbit|0⟩} c:Circuit(qbit⊗qbit⊗qbit, qbit) {b↦qbit\kp} =
		box (q, a, b) ⇒
		(* 1. Bell state preparation for Alice and Bob *)
		a       ← H a;
		(a, b)  ← CX (a, b);
		(* 2. Bell measurement by Alice *)
		(q, a)  ← CX (q, a);
		q       ← H q;
		x       ← meas q;
		y       ← meas a;
		(* 3. Correction by Bob *)
		(y, b)  ← CX (y, b);
		(x, b)  ← CZ (x, b);
		output b
		\end{lstlisting}
		% ()      ← discard y;
		%     ()      ← discard x;
	\end{subfigure}
	\begin{subfigure}{1in}
		% 	\centerline{
		\Qcircuit @C=.4em @R=.4em @! {
			\lstick{q=\ket{\psi}} & \qw & \qw & \ctrl{1} & \gate{H} & \meter & \cw & \control \cw\\
			\lstick{a=\ket{0}} & \gate{H} & \ctrl{1} & \gate{X} & \qw & \meter & \control \cw & \\
			\lstick{b=\ket{0}} & \qw & \gate{X} & \qw & \qw & \qw & \gate{X} \cwx & \gate{Z} \cwx[-2] & \rstick{\ket{\psi}} \qw
			\gategroup{2}{2}{3}{3}{.7em}{.}
			\gategroup{1}{4}{2}{6}{.7em}{.}
		}
		% }
	\end{subfigure}
	\Description{On the left, \qwire code to implement Quantum Teleportation algorithm annotated by its Hoare Type that provides specification for the algorithm. On the right, the corresponding quantum circuit.}
	\caption{Quantum Teleportation (Code and Circuit)}
	\label{fig:teleport}
\end{figure}

In Fig. \ref{fig:teleport} we show a function \lstinline{teleport} that implements the quantum teleportation circuit. It returns one qubit and takes three qubits as arguments: one qubit \lstinline{q} that could be in any unknown state $\ket{\psi}$ which Alice is trying to send to Bob; and two qubits, \lstinline{a} and \lstinline{b}, initially in the state $\ket{0}$. The function first entangles the latter two qubits into a Bell state. Then, Alice performs a Bell measurement of her two qubits \lstinline{q} and \lstinline{a} and sends the resulting classical bits (\lstinline{x} and \lstinline{y}) to Bob. Finally, Bob applies corrective measurement using the quantum gates X and Z controlled by the classical bits that he received from Alice. The result is that Bob's qubit is now in the same state as the initial state of the message qubit \lstinline{q}. The specification that the function must meet is written using a Hoare type, which we describe next.

@@

\subsection{Verification}
Here we describe how HTT lets us verify the correctness of teleportation based on the specification provided as the type of the function. Type checking in HTT involves generation of strongest postconditions at each step of the program with the initial precondition as input. Verification is a two-step process: the first phase does basic type checking and verification condition generation which is decidable, the second phase needs to show the validity of the generated verification conditions, which is undecidable. This second phase can be deferred to an automated theorem prover. For illustration, we show the assertions as we step through step 1 of teleportation. The final state here is that qubits \lstinline{a} and \lstinline{b} are entangled into the first Bell state, $\ket{\phi^+} = \frac{1}{\sqrt{2}}(\ket{00} + \ket{11})$.

@@

\subsection{Programming Languages for Quantum Computing}
\textit{A chronology of langugage developments\ldots}

@@

Our approach builds upon previous work in reasoning about quantum programs such as Quantum Weakest Preconditions~\cite{dhondt_quantum_2006} and Quantum Hoare Logic~\cite{ying_floydhoare_2012} in the spirit of Hoare~\cite{hoare_axiomatic_1969} and Dijkstra~\cite{dijkstra_discipline_1976} but attempts to bring those reasoning techniques into the type system. The hope is that programmers will be able to encode some of the semantic properties that they expect of their programs as specifications in their code and type checking will ensure correctness of some of those properties. In the classical setting, Hoare Type Theory (HTT)~\cite{nanevski_hoare_2008} accomplishes exactly this goal. Here, we attempt to merge these ideas for the quantum realm.

@@

We also look to explore how our types will evolve when we start incorporating higher-order constructs in our quantum language such as those demonstrated in languages like Quantum $\lambda$-calculus~\cite{selinger_lambda_2006} and Quipper~\cite{green_quipper:_2013}.

>>>>

\documentclass[acmsmall,nonacm,timestamp]{acmart}

@@

\begin{abstract}
Quantum Computing is a new and upcoming way to utilize quantum mechanics for computation. It is easy to introduce bugs while programming quantum computers similar to classical computing. Strong static type systems have prevented a lot of bugs from being introduced in the classical paradigm. There is a similar need for well-designed static types in the domain of quantum computing. We propose Quantum Hoare Types that let programmers reason statically about certain semantic properties of their quantum programs and further allow a limited form of formal verification.
\end{abstract}

@@

\section{Introduction}
It is difficult to reason about the correctness of quantum programs. Sound static type systems help prevent a huge class of bugs from occurring but since the realm of quantum programming is still new there is not a lot of consensus on what kind of types make the most sense. Further, it is unclear how much they help programmers reason about the semantic properties associated with the quantum algorithms that they are implementing.

@@

\section{Background}

\subsection{Quantum Computing}
The basic unit of information in quantum computing is a qubit, that can be represented using a unit vector in a two-dimensional complex vector space. Using Dirac's bra-ket notation, we can write an arbitrary qubit, $\ket{\psi}$, as a linear combination of the computational basis vectors: $\ket{\psi} = \alpha\ket{0} + \beta\ket{1}$, where $\ket{0}=\begin{bsmallmatrix}
1 \\
0
\end{bsmallmatrix}$ and $\ket{1}=\begin{bsmallmatrix}
0 \\
1
\end{bsmallmatrix}$, and the two constants are complex amplitudes that are normalized using the constraint $|\alpha|^2 +|\beta|^2 = 1$. This means that when measured, the probability of the qubit being in the state 0 is $|\alpha|^2$ and that for the state 1 is $|\beta|^2$. To work with multiple qubits, we need to take their tensor product, denoted by $\otimes$. Computation is performed by the application of unitary quantum gates, such as the Hadamard (H) gate or the controlled NOT (CX) gate, that can be represented as square matrices. Mathematically, a gate application is equivalent to multiplying the matrix representing the gate with the state vector of the qubit(s).

This basic unit of information is not the only reason why quantum computing is fundamentally different from classical computing---certain quantum mechanical phenomenon are highly counterintuitive! One such phenomenon is \textit{quantum entanglement} which, at its most basic level, involves a correlation between two qubits such that when one is measured, the outcome obtained necessarily influences the measurement outcome of the other qubit (even if the two qubits are far apart). We will see how this can be utilized for computation in the quantum teleportation algorithm that we describe and program in \S\ref{telep}. We refer the reader to the standard textbook~\cite{nielsen_quantum_2010} or Nielsen's excellent series of essays~\cite{matuschak2019} for more background.

@@

\subsection{Example: Quantum Teleportation}\label{telep}
Teleportation is a simple quantum algorithm that is used to communicate a quantum state using two classical bits and a shared entangled pair of qubits. Alice is trying to send a message encoded in the qubit to Bob. There are three broad steps in the protocol: 1) preparation, ie, entanglement creation between two qubits to be shared between Alice and Bob in advance; 2) sending the message by Alice; and, 3) reconstruction of the message by Bob.
\begin{figure}[h]
	\begin{subfigure}{3in}
		% 	a       ← init0 ();
		%     b       ← init0 ();
		\begin{lstlisting}
		teleport : Π\kp:qbit.{q↦qbit\kp /\ a↦qbit|0⟩ /\ b↦qbit|0⟩} c:Circuit(qbit⊗qbit⊗qbit, qbit) {b↦qbit\kp} =
		box (q, a, b) ⇒
		(* 1. Bell state preparation for Alice and Bob *)
		a       ← H a;
		(a, b)  ← CX (a, b);
		(* 2. Bell measurement by Alice *)
		(q, a)  ← CX (q, a);
		q       ← H q;
		x       ← meas q;
		y       ← meas a;
		(* 3. Correction by Bob *)
		(y, b)  ← CX (y, b);
		(x, b)  ← CZ (x, b);
		output b
		\end{lstlisting}
		% ()      ← discard y;
		%     ()      ← discard x;
	\end{subfigure}
	\begin{subfigure}{1in}
		% 	\centerline{
		\Qcircuit @C=.4em @R=.4em @! {
			\lstick{q=\ket{\psi}} & \qw & \qw & \ctrl{1} & \gate{H} & \meter & \cw & \control \cw\\
			\lstick{a=\ket{0}} & \gate{H} & \ctrl{1} & \gate{X} & \qw & \meter & \control \cw & \\
			\lstick{b=\ket{0}} & \qw & \gate{X} & \qw & \qw & \qw & \gate{X} \cwx & \gate{Z} \cwx[-2] & \rstick{\ket{\psi}} \qw
			\gategroup{2}{2}{3}{3}{.7em}{.}
			\gategroup{1}{4}{2}{6}{.7em}{.}
		}
		% }
	\end{subfigure}
	\Description{On the left, \qwire code to implement Quantum Teleportation algorithm annotated by its Hoare Type that provides specification for the algorithm. On the right, the corresponding quantum circuit.}
	\caption{Quantum Teleportation (Code and Circuit)}
	\label{fig:teleport}
\end{figure}

In Fig. \ref{fig:teleport}, on the left, we show a function \lstinline{teleport} written in \qwire-like syntax (TODO: add ref) that implements the quantum teleportation circuit shown on the right. It returns one qubit and takes three qubits as arguments: one qubit \lstinline{q} that could be in any unknown state $\ket{\psi}$ which Alice is trying to send to Bob; and two qubits, \lstinline{a} and \lstinline{b}, initially in the state $\ket{0}$. The function first entangles the latter two qubits into a Bell state. Then, Alice performs a Bell measurement of her two qubits \lstinline{q} and \lstinline{a} and sends the resulting classical bits (\lstinline{x} and \lstinline{y}) to Bob. Finally, Bob applies corrective measurement using the quantum gates X and Z controlled by the classical bits that he received from Alice. The result is that Bob's qubit is now in the same state as the initial state of the message qubit \lstinline{q}. The specification that the function must meet is written using a Hoare type, which we describe next.

@@

\subsection{Verification}
Here we describe how HTT lets us verify the correctness of teleportation based on the specification provided as the type of the function. Type checking in HTT involves generation of strongest postconditions at each step of the program with the initial precondition as input. Verification is a two-step process: the first phase does basic type checking and verification condition generation which is decidable, the second phase needs to show the validity of the generated verification conditions, which is undecidable. This second phase can be deferred to an automated theorem prover. For illustration, we show the assertions as we step through the first step of teleportation. The final state here is that qubits \lstinline{a} and \lstinline{b} are entangled into the first Bell state, $\ket{\phi^+} = \frac{1}{\sqrt{2}}(\ket{00} + \ket{11})$.

@@

\subsection{Programming Languages for Quantum Computing}
\textit{A chronology of language developments\ldots}

@@

Our approach builds upon previous work in reasoning about quantum programs such as Quantum Weakest Preconditions~\cite{dhondt_quantum_2006} and Quantum Hoare Logic~\cite{ying_floydhoare_2012} in the spirit of \citet{hoare_axiomatic_1969} and \citet{dijkstra_discipline_1976} but attempts to bring those reasoning techniques into the type system. The hope is that programmers will be able to encode some of the semantic properties that they expect of their programs as specifications in their code and type checking will ensure correctness of some of those properties. In the classical setting, Hoare Type Theory (HTT)~\cite{nanevski_hoare_2008} accomplishes exactly this goal. Here, we attempt to merge these ideas for the quantum realm.

@@

We also look to explore how our types will evolve when we start incorporating higher-order constructs in our quantum language such as those demonstrated in languages like quantum lambda calculus~\cite{selinger_lambda_2006} and Quipper~\cite{green_quipper:_2013}.
